Multi-target detection with application to cryo-electron microscopy

### 2019-02-15 inverse_problems/inverse_problems.tex
Intro done

\subsection{Problem formulation} \label{sec:problem_formulation}

We consider the \emph{multi-target detection} problem of recovering a set of $K$ signals that appear 
multiple times at unknown locations in a noisy measurement.
Let $x_1,\ldots,x_K\in\RL$ be the sought signals and let $y\in\RN$ be the observed data, where we assume $N$ is  far larger than $L$. 
Let  $s[i]$ count the number of signals whose first element is positioned in $y[i]$, each of those  is chosen according to some (possibly unknown) distribution over $\{1,\ldots,K\}$. 
If signal occurrences overlap, they interfere additively. %We think of those events as being rare. 
With additive white Gaussian noise, the measurement model can be written as 
\begin{equation} 
y  =  \sum_{k=1}^K s_k \ast x_k + \varepsilon, \qquad  \varepsilon   \sim \mathcal{N}(0,\sigma^2 I_N),
\label{eq:model}
\end{equation}
where $\ast$ denotes linear convolution, and $s_k[i]$ indicates the number of starting positions of $x_k$ in $y[i]$ so that $s =  s_1+\cdots+s_K$. Explicitly, with zero-based indexing, % and zero-padding out of bounds,
\begin{align*}
	y[i] & = \sum_{k=1}^{K} \sum_{j = 0}^{L-1} s_k[i-j] x_k[j] + \varepsilon[i].
\end{align*}
The goal is  to estimate $x_1,\ldots,x_K$ from $y$. % while $s$ is the \emph{nuisance variable} of the problem.
%We name this problem \emph{multi-target detection}.
This idealized setup appears in several scientific applications, including structural biology~\cite{bendory2018toward} (as we detail below), spike sorting~\cite{lewicki1998review}, passive radar~\cite{gogineni2017passive}, and system identification~\cite{ljung1998system}.

@@

\subsection{Models for target distribution}

We consider two  models for the signal $s$: the first one is .......

@@

As a first setup, we allow any generative model for $s$ which meets the following separation requirement: $s$ is binary, and
\begin{equation}
\textrm{If } s[i] = 1 \textrm{ and } s[j] = 1 \textrm{ for } i \neq j, \textrm{ then } |i - j| \geq 2L-1.
\label{eq:spacing}
\end{equation}
In words: the starting positions of any two occurrences  must be separated by at least $2L-1$ positions, so that their end points are necessarily separated by at least $L-1$ signal-free entries in the data.
%Under the separation condition~\eqref{eq:spacing}, we refer to this model as the \emph{homogeneous well-separated model}. 
%In the next section we present a model that alleviates this condition.
%As will be shown next, in a low SNR environment, estimating $s$ is impossible, whereas in some cases estimating $x$ is tractable.
This property ensures that correlating $y$ with versions of itself shifted by at most $L-1$ entries does not involve correlating distinct signal occurrences. As a result, the only properties of $s_1, \ldots, s_K$ that affect the autocorrelations of $y$ are the total number of occurrences of the distinct signals: their individual and relative locations do not intervene.

@@

Specifically, for some parameter $\gamma > 0$, the total number of signal occurrences is drawn from a Poisson distribution with expectation $(N-L+1)\gamma$. For each occurrence, a starting location $i$ is drawn uniformly and independently at random from $\{0, \ldots, N-L\}$. Then, one of the signals $x_1, \ldots, x_K$ is selected independently at random from a fixed distribution and added to the observation $y$ with first entry positioned at $i$. 
By definition of the Poisson process, the entries of $s$ are independent and follow a Poisson distribution with expectation $\gamma$. In particular, this model allows for (additive) signal overlap. In Section~\TODO{ref}, we show that the autocorrelations of $y$ are nonetheless equivalent to those arising under our well-separated model above.

\subsection{Extensions} \label{sec:extensions}

\TODO{multi-dimensions}

The basic model~\eqref{eq:model} can be extended in various ways, some discussed below.
%
%\paragraph{Multi-target detection and clustering with separation.}
%The multi-target detection model can be naturally generalized to the problem of estimating a set of $K$ signals $x_1,\ldots,x_K\in\RL$. Let $s$ be a binary signal satisfying the separation condition~\eqref{eq:spacing}.
%For each $s[i]=1$, a random variable $v_i$ is drawn from some (possibly unknown) distribution over $\{1,\ldots,K\}$. Then, $x_{v_i}$ is placed in $y$
%with element $0$ at location $i$. 
%We refer to this model as \emph{multi-target detection and clustering}.
%
% This model can be written as a mix of blind deconvolution problems
%\begin{equation} 
%y  =  \sum_{k=1}^Kx_k \ast s_k + \varepsilon, \quad  \varepsilon   \sim \mathcal{N}(0,\sigma^2 I_N),
%\end{equation}
%where again $s_i$ is a binary signal, which its non-zero values 
%are the subset of the non-zero values of $s$ associated with $x_i$.
%Generally, we refer to this problem as \emph{multi-target detection and clustering with separation.} 
%An key question in this model is how many signals can be recovered simultaneously. We derive a bound in Section~\ref{sec:heterogeneity} and show a numerical experiment in Section~\ref{sec:numerics}.

\paragraph{The well-separated model with a random signal}
The homogeneous and heterogeneous models can be generalized as follows. 
We let the signal $X\in\RL$ to be a random vector drawn from some fixed distribution. %We observe a random vector $Y\in\R^{N-L+1}$.
For each non-zero element of $s$,  a random vector $X$ from the distribution is then placed in $y$, with element $0$ at location $i$.
If the distribution is a Delta function that this model coincides with the multi-target detection model; if it is a sum of Diracs then it reduces to multi-target detection and clustering.

>>>>

%\subsection{Problem formulation} \label{sec:problem_formulation}

We consider the \emph{multi-target detection} problem of recovering a set of $K$ signals that appear 
multiple times at unknown locations in a noisy measurement.
Let ${x_1,\ldots,x_K\in\RL}$ be the sought signals and let $y\in\RN$ be the observed data, where we assume $N$ is  far larger than $L$. 
Let  $s[i]$ count the number of signals whose first element is positioned in $y[i]$. Each of those $s[i]$ is chosen according to some (possibly unknown) distribution $\pi$ \TODO{Add notation for the distribution} over $\{1,\ldots,K\}$. 
If signal occurrences overlap, they interfere additively. %We think of those events as being rare. 
With additive white Gaussian noise, the measurement model can be written as 
\begin{equation} 
y  =  \sum_{k=1}^K s_k \ast x_k + \varepsilon, \qquad  \varepsilon   \sim \mathcal{N}(0,\sigma^2 I_N),
\label{eq:model}
\end{equation}
where $\ast$ denotes linear convolution, and $s_k[i]$ indicates the number of starting positions of $x_k$ in $y[i]$ so that $s =  s_1+\cdots+s_K$. Explicitly, with zero-based indexing, % and zero-padding out of bounds,
\begin{align*}
	y[i] & = \sum_{k=1}^{K} \sum_{j = 0}^{L-1} s_k[i-j] x_k[j] + \varepsilon[i].
\end{align*}
The goal is  to estimate $x_1,\ldots,x_K$ from $y$. % while $s$ is the \emph{nuisance variable} of the problem.
%We name this problem \emph{multi-target detection}.
This idealized setup appears in several scientific applications, including structural biology~\cite{bendory2018toward} (as we detail below), spike sorting~\cite{lewicki1998review}, passive radar~\cite{gogineni2017passive}, and system identification~\cite{ljung1998system}.

@@

\subsection*{Models for target distribution}

We consider two  models for the distribution of signal occurrences in the observation, that is, for $s_1, \ldots, s_K$.

@@

As a first setup, we allow any generative model for $s$ which meets the following separation requirement: $s$ is binary, and
\begin{equation}
\textrm{If } s[i] = 1 \textrm{ and } s[j] = 1 \textrm{ for } i \neq j, \textrm{ then } |i - j| \geq 2L-1.
\label{eq:spacing}
\end{equation}
In words: the starting positions of any two occurrences  must be separated by at least $2L-1$ positions, so that their end points are necessarily separated by at least $L-1$ signal-free entries in the data.
%Under the separation condition~\eqref{eq:spacing}, we refer to this model as the \emph{homogeneous well-separated model}. 
%In the next section we present a model that alleviates this condition.
%As will be shown next, in a low SNR environment, estimating $s$ is impossible, whereas in some cases estimating $x$ is tractable.
This property ensures that correlating $y$ with versions of itself shifted by at most $L-1$ entries does not involve correlating distinct signal occurrences. Once $s$ is determined, for each position $i$ such that $s[i] = 1$, one of the signals $x_k$ is selected independently at random, and accordingly we set $s_k[i] = 1$. As a result, the only properties of $s_1, \ldots, s_K$ that affect the autocorrelations of $y$ are the total number of occurrences of the distinct signals: their individual and relative locations do not intervene.

@@

Specifically, for some parameter $\gamma > 0$, the total number of signal occurrences is drawn from a Poisson distribution with expectation $(N-L+1)\gamma$. For each occurrence, a starting location $i$ is drawn uniformly and independently at random from $\{0, \ldots, N-L\}$. Then, one of the signals $x_1, \ldots, x_K$ is selected independently at random from a fixed distribution and added to the observation $y$ with first entry positioned at $i$. 
By definition of the Poisson process, the entries of $s$ are independent and follow a Poisson distribution with expectation $\gamma$. In particular, this model allows for (additive) signal overlap. In Section~\ref{sec:poisson}, we show that the autocorrelations of $y$ are nonetheless equivalent to those arising under our well-separated model above.

\subsection*{Extensions} \label{sec:extensions}

Extending the problem setup and autocorrelation analysis to signals in more than one dimension is straightforward: see discussion in Section~\ref{sec:high_dimensions} and numerical experiments in Section~\ref{sec:numerics}.

Likewise, it is easy to extend the model to situations where the signal occurrences are sampled from a general distribution rather than from a finite set of choices $x_1, \ldots, x_K$. In this setup, the goal is to estimate the distribution (possibly defined by a finite set of parameters). In particular, this allows for continuous distributions of targets. We adopt this perspective when deriving the autocorrelations in Section~\ref{sec:AC_analysis}.

In the next section, we show how this flexibility allows us to model an important imaging problem in structural biology.
